Convective heat transport in stratified atmospheres at low and high Mach number

### 2017-06-19 referee_process/second_submission/referee_response.tex

\begin{myquotation}
This Letter reports results from a numerical study of thermal
convection in a compressible fluid at unity Prandtl number and a range
of initial stratification. The main result is the scaling of the
Nusselt number with the Rayleigh number, supplemented by some
observations such as characteristic changes in the transition between
subsonic and supersonic regimes.

This problem is highly challenging. Ideally, a numerical exploration
should be guided by some theoretical insights or practical
observations. The present paper appears to be ``purely'' numerical
without such guidance, albeit some comparison with the well-known
problem of incompressible Rayleigh--Bernard convection. Nevertheless,
the results are significant and should be published. While I do not
wholeheartedly recommend this paper to PRL, I would not object to its
acceptance.

Below is a minor comment that the authors may find useful.

Most parts of the paper is relatively well written, except in the
introduction. Here, there is room for improvement. The introductory
paragraph is virtually empty. I would cite a couple of papers for
"These prior studies" and be specific about both "important insight"
and known "fundamental properties". The final paragraph of the
introduction has failed to fulfill its mission. I would explicitly
state (at least) the most significant result of the paper in this
paragraph.
\end{myquotation}
We have clarified the structure of the introduction paragraph, have included
citations to prior work at high Ma, and have set the context for the study
of fundamental heat transport conducted here at low as well as high Ma.

### 2017-06-19 tex/nusselt_paper.tex
work with Ben

\begin{abstract}
%Convection in astrophysical systems is stratified and
%often occurs at high Rayleigh number (Ra) and low
%Mach number (Ma).
We study stratified convection in the context of 
plane-parallel, polytropically stratified atmospheres. 
We perform a suite of 2D and 3D simulations in which we vary the initial
superadiabaticity ($\epsilon$) and the Rayleigh number (Ra) while fixing the
initial density stratification, aspect
ratio, and Prandtl number.
The evolved heat transport, 
quantified by the Nusselt number (Nu),
follows scaling relationships similar to those found in the well-studied, 
incompressible Rayleigh-B\'{e}nard problem in both 2D and 3D and is not
appreciably affected by the magnitude of $\epsilon$.
The evolved Mach number (Ma) scales according to
$\text{Ma} \propto \epsilon^{1/2}\text{Ra}^{\alpha}$, where
$\alpha = 0$ in 3D and $\alpha \approx 1/4$ in 2D.
At large values of $\epsilon$, significant density inversions appear in the
evolved atmospheres and seem to persist in both 2D and 3D.
\end{abstract}
\maketitle

@@

In RB convection, there exist two primary dynamical control parameters: 
the Rayleigh number (Ra, the ratio of
buoyant driving to diffusive damping) and the Prandtl number 
(Pr, the ratio of viscous to thermal
diffusivity). These numbers control two useful
measures of turbulence in the evolved solution:
the Reynolds
number (Re, the strength of advection to viscous diffusion)
and the Peclet number (Pe, advection vs. thermal diffusion).  
In stratified atmospheres, the magnitude of the unstable entropy gradient
joins Ra and Pr as an important control parameter.  This 
\emph{superadiabatic excess} \cite{graham1975}, $\epsilon$,
sets the scale of the atmospheric entropy gradient.
We find here that $\epsilon$ primarily controls the Ma of the evolved solution.

@@

\section{Experiment} 
\refstepcounter{section}
\label{sec:experiment}
We examine a monatomic ideal gas with an adiabatic index of
$\gamma = 5/3$ whose equation of state is $P = R\rho T$. This is consistent with the approach used in earlier work 
\cite{graham1975, chan&all1982, brandenburg&all2005,
hurlburt&all1984, cattaneo&all1990, cattaneo&all1991, brummell&all1996} 
and is the simplest stratified extension of RB.
We study atmospheres which are initially polytropically stratified,
\begin{equation}
\begin{split}
\rho_0(z) &= \rho_{t}(1 + L_z - z)^m, \\
T_0(z)    &= T_{t}(1 + L_z - z),
\label{eqn:polytrope}
\end{split}
\end{equation}
where $m$ is the polytropic index and $L_z$ is the depth of the atmosphere.
The polytropic
index is set by the superadiabatic excess, $\epsilon = m_{ad} - m$, where
$m_{ad} = (\gamma - 1)^{-1}$ is the adiabatic value of $m$.
The height coordinate, $z$, increases upwards in the range $[0, L_z]$.
Subscript 0 indicates initial conditions and subscript $t$ indicates values
at $z = L_z$.   We
specify the depth of the atmosphere, $L_z = e^{n_{\rho}/m} - 1$, by choosing
the number of density scale heights, $n_{\rho}$, it spans initially.
Throughout this letter we set $n_{\rho} = 3$.    Satisfying hydrostatic
equilibrium sets the value of gravity, $g = RT_t (m + 1)$, which is
constant with depth.  We study atmospheres with aspect
ratios of 4 where both the $x$ and $y$ coordinates have the range $[0, 4L_z]$.
In our 2D cases, we only consider $x$ and $z$.

@@

At fixed $n_\rho$, convective dynamics are 
controlled by $\epsilon$ as well as the atmospheric diffusivities.
At a fixed value of
$\epsilon$, the diffusivities are set by the
Rayleigh number (Ra) and the Prandtl number (Pr),
\begin{equation}
\text{Ra}_{t} = \frac{g L_z^3 (\Delta S_0 / c_P)}{\nu_t\chi_t},
\qquad
\text{Pr} = \frac{\nu}{\chi},
\end{equation}
where $\Delta S_0 = \epsilon\ln (1 + L_z)$ is the initial entropy
difference between the top and bottom boundaries,
$c_P = \gamma/(\gamma-1)$ is the specific heat a
constant pressure, and $\nu$ and $\chi$ are the thermal diffusivity and
kinematic viscosity, respectively.
Throughout this work we specify
that Pr $= 1$ and is depth invariant.
Polytropes, as specified in (\ref{eqn:polytrope}),
are in thermal equilibrium. Thus, the initial thermal
conductivity, $\kappa_0 = \chi \rho_0$, is
constant with depth. 
By these
choices, $\nu(z) \equiv \chi(z) \equiv \chi_t / \rho_0$.
This formulation 
sets Ra at the bottom of the domain greater than
Ra$_t$ by a factor of $e^{2n_\rho}$. Henceforth
when we specify Ra we are referring to Ra$_t$.  
The full values of $\kappa = \rho\chi$ and 
$\mu = \rho\nu$ (the dynamic viscosity) are free to evolve as the density 
profile evolves.

Under this formulation, diffusivities scale as
\begin{equation}
\chi_t = \nu_t \propto \sqrt{\frac{\epsilon}{\text{Ra}_t}}
\label{eqn:diffusivities}
\end{equation}
for constant Pr = 1, as in this work.
We carry out two experiments in this letter: one in which
we fix $\epsilon$ and increase Ra, thus increasing the diffusive
timescales, and a second in which we fix Ra and raise $\epsilon$,
thus modifying the ratio of the diffusive timescales to the
buoyant timescale.

@@

Low Ma flows (e.g., $\epsilon = 10^{-4}$)
display the classic narrow downflow and broad upflow lanes of stratified
convection (Fig. \ref{fig:entropy_snapshots}a).
At high Ma (e.g., $\epsilon = 0.5$, Ra/Ra$_{\text{crit}} \gtrsim 10^3$), 
bulk thermodynamic structures are similar but
shock systems form in the upper atmosphere near downflow lanes 
(Fig. \ref{fig:entropy_snapshots}b\&c).
These shock phenomena were reported in
two \cite{cattaneo&all1990} and 
three \cite{malagoli&all1990} dimensional polytropic simulations previously.
As in (\ref{eqn:diffusivities}), diffusivities become small as Ra 
is increased to large values (Fig. \ref{fig:entropy_snapshots}c) and
thermodynamic structures break up into small eddies which traverse the
domain repeatedly before diffusing.
Furthermore, the value of $\epsilon$ sets the size of the
evolved thermodynamic fluctuations from their adiabatic values,
as evidenced by the scalings of the colorbars.
%These fluctuations are very small when $\epsilon$ and Ma are small,
%but can be O(1) for values of $\epsilon$ near 1.

@@

The rms Reynolds number (Re = $|\bm{u}|L_z/\nu$) and Peclet number
(Pe = Pr Re)
compare the importance of advection to diffusion in the evolved
convective state.  For Pr = 1, such as in this work, Pe = Re.  
Our choice of $\{\nu,\chi\}\propto \rho_0^{-1}$ drastically changes
the value of Re between the top and bottom of the atmosphere.  We report values of
Re at the midplane ($z=L_z/2$) of the atmosphere in
Fig. \ref{fig:re_and_nu_v_ra}b.  Largely we find that Re
depends on Ra, but not $\epsilon$, except when the flow states
change.
In 2D and at low Ra, Re $\appropto$ Ra$^{3/4}$.  The heightened scaling
of Re in 2D is due to the scaling of velocity (Ma) with Ra, as is
seen in Fig. \ref{fig:ma_v_eps}.  When the flows
become supersonic, 
this scaling gives way to Re $\appropto$ Ra$^{1/2}$, as expected
by the scaling of diffusivities in (\ref{eqn:diffusivities}).
In our limited 3D runs, the scaling of Re is
Re $\appropto$ Ra$^{1/2}$, consistent with the supersonic results.

>>>>

\begin{abstract}
%Convection in astrophysical systems is stratified and
%often occurs at high Rayleigh number (Ra) and low
%Mach number (Ma).
We study stratified convection in the context of 
plane-parallel, polytropically stratified atmospheres. 
We perform a suite of 2D and 3D simulations in which we vary the initial
superadiabaticity ($\epsilon$) and the Rayleigh number (Ra) while fixing the
initial density stratification, aspect
ratio, and Prandtl number.
The evolved heat transport, 
quantified by the Nusselt number (Nu),
follows scaling relationships similar to those found in the well-studied, 
incompressible Rayleigh-B\'{e}nard problem.  This scaling holds up in both 2D and 3D and is not
appreciably affected by the magnitude of $\epsilon$. %think about last sentence.
\end{abstract}
\maketitle

@@

In RB convection, there exist two primary dynamical control parameters: 
the Rayleigh number (Ra, the ratio of
buoyant driving to diffusive damping) and the Prandtl number 
(Pr, the ratio of viscous to thermal
diffusivity). These numbers control two useful
measures of turbulence in the evolved solution:
the Reynolds
number (Re, the strength of advection to viscous diffusion)
and the Peclet number (Pe, advection vs. thermal diffusion).  
In stratified atmospheres, the magnitude of the unstable entropy gradient
joins Ra and Pr as a third important and independent control parameter.  This 
\emph{superadiabatic excess} \cite{graham1975}, $\epsilon$,
sets the scale of the atmospheric entropy gradient.
We find here that $\epsilon$ primarily controls the Ma of the evolved solution.

@@

\section{Experiment} 
\refstepcounter{section}
\label{sec:experiment}
We examine a monatomic ideal gas with an adiabatic index of
$\gamma = 5/3$ whose equation of state is $P = R\rho T$. This is consistent with the approach used in earlier work 
\cite{graham1975, chan&all1982, brandenburg&all2005,
hurlburt&all1984, cattaneo&all1990, cattaneo&all1991, brummell&all1996} 
and is the simplest stratified extension of RB.
We study atmospheres which are initially polytropically stratified,
\begin{equation}
\begin{split}
\rho_0(z) &= \rho_{t}(1 + L_z - z)^m, \\
T_0(z)    &= T_{t}(1 + L_z - z),
\label{eqn:polytrope}
\end{split}
\end{equation}
where $m$ is the polytropic index and $L_z$ is the depth of the atmosphere.
The polytropic
index is set by the superadiabatic excess, $\epsilon = m_{ad} - m$, where
$m_{ad} = (\gamma - 1)^{-1}$ is the adiabatic value of $m$.
The height coordinate, $z$, increases upwards in the range $[0, L_z]$.
Subscript 0 indicates initial conditions and subscript $t$ indicates values
at $z = L_z$.   Stratified atmospheres have a fourth non-dimensional parameter,
the number of density scale heights, $n_{\rho} = \ln\left[\rho_0(z=0)/\rho_t\right]$.  We
specify the depth of the atmosphere, $L_z = e^{n_{\rho}/m} - 1$, by choosing
the initial value of $n_{\rho}$.
Throughout this letter we set $n_{\rho} = 3$.    Satisfying hydrostatic
equilibrium sets the value of gravity, $g = RT_t (m + 1)$, which is
constant with depth.  We study atmospheres with aspect
ratios of 4 where both the $x$ and $y$ coordinates have the range $[0, 4L_z]$.
In our 2D cases, we only consider $x$ and $z$.

@@

At fixed $n_\rho$, convective dynamics are 
controlled by $\epsilon$ as well as the atmospheric diffusivities.
At a fixed value of
$\epsilon$, the diffusivities are set by the
Rayleigh number (Ra) and the Prandtl number (Pr),
\begin{equation}
\text{Ra}_{t} = \frac{g L_z^3 (\Delta S_0 / c_P)}{\nu_t\chi_t},
\qquad
\text{Pr} = \frac{\nu}{\chi},
\end{equation}
where $\Delta S_0 = \epsilon\ln (1 + L_z)$ is the initial entropy
difference between the top and bottom boundaries,
$c_P = \gamma/(\gamma-1)$ is the specific heat a
constant pressure, and $\nu$ and $\chi$ are the thermal diffusivity and
kinematic viscosity, respectively.
Throughout this work we specify
that Pr $= 1$ and is depth invariant.
Polytropes, as specified in (\ref{eqn:polytrope}),
are in thermal equilibrium. Thus, the initial thermal
conductivity, $\kappa_0 = \chi \rho_0$, is
constant with depth. 
By these
choices, $\nu(z) \equiv \chi(z) \equiv \chi_t / \rho_0$.
This formulation 
sets Ra at the bottom of the domain greater than
Ra$_t$ by a factor of $e^{2n_\rho}$. Henceforth
when we specify Ra we are referring to Ra$_t$.  
The full values of $\kappa = \rho\chi$ and 
$\mu = \rho\nu$ (the dynamic viscosity) are free to evolve as the density 
profile evolves.  
Under this formulation, diffusivities scale as
$\chi_t, \nu_t \propto \sqrt{g L_z^3 \epsilon / \text{Ra}_t}$.
The thermal diffusion timescale is $t_\chi \equiv L_z^2 / \chi$. The
ratio of the diffusion time to the buoyancy time is
\begin{equation}
\frac{t_\chi}{t_b} = \left(\frac{c_P}{n_\rho}\right)^{1/2}\text{Ra}_t^{1/2}
\left(m_{ad} - \epsilon\right)^{1/2}.
\label{eqn:timescales}
\end{equation}
We carry out two experiments in this letter: one in which
we fix $\epsilon$ and increase Ra, thus increasing the ratio in
(\ref{eqn:timescales}), and a second in which we fix Ra and change $\epsilon$.
In this second experiment, the buoyancy time $t_b$ scales as $\epsilon^{-1/2}$
relative to the speed of sound, and we see this reflected in the Mach
number scaling. The ratio in (\ref{eqn:timescales}) is
independent of $\epsilon$ when $\epsilon \ll m_{ad}$.

@@

Low Ma flows (e.g., $\epsilon = 10^{-4}$)
display the classic narrow downflow and broad upflow lanes of stratified
convection (Fig. \ref{fig:entropy_snapshots}a).
At high Ma (e.g., $\epsilon = 0.5$, Ra/Ra$_{\text{crit}} \gtrsim 10^3$), 
bulk thermodynamic structures are similar but
shock systems form in the upper atmosphere near downflow lanes 
(Fig. \ref{fig:entropy_snapshots}b\&c).
These shock phenomena were reported in
two \cite{cattaneo&all1990} and 
three \cite{malagoli&all1990} dimensional polytropic simulations previously.
As in (\ref{eqn:timescales}), diffusivities become small as Ra 
is increased to large values (Fig. \ref{fig:entropy_snapshots}c) and
thermodynamic structures break up into small eddies which traverse the
domain repeatedly before diffusing.
Furthermore, the value of $\epsilon$ sets the size of the
evolved thermodynamic fluctuations from their adiabatic values,
as evidenced by the scalings of the colorbars.
%These fluctuations are very small when $\epsilon$ and Ma are small,
%but can be O(1) for values of $\epsilon$ near 1.

@@

The rms Reynolds number (Re = $|\bm{u}|L_z/\nu$) and Peclet number
(Pe = Pr Re)
compare the importance of advection to diffusion in the evolved
convective state.  For Pr = 1, such as in this work, Pe = Re.  
Our choice of $\{\nu,\chi\}\propto \rho_0^{-1}$ drastically changes
the value of Re between the top and bottom of the atmosphere.  We report values of
Re at the midplane ($z=L_z/2$) of the atmosphere in
Fig. \ref{fig:re_and_nu_v_ra}b.  Largely we find that Re
depends on Ra, but not $\epsilon$, except when the flow states
change.
In 2D and at low Ra, Re $\appropto$ Ra$^{3/4}$.  The heightened scaling
of Re in 2D is due to the scaling of velocity (Ma) with Ra, as is
seen in Fig. \ref{fig:ma_v_eps}.  When the flows
become supersonic, 
this scaling gives way to Re $\appropto$ Ra$^{1/2}$, as expected
by the scaling of diffusivities in (\ref{eqn:timescales}).
In our limited 3D runs, the scaling of Re is
Re $\appropto$ Ra$^{1/2}$, consistent with the supersonic results.
